A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-07 rigid_bodies.tex
Almost finished the NVT integration section

\subsection{Integration Schemes for the Canonical Ensemble}

As done in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion. First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components. In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator}. Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as

\begin{equation}
\label{eq:trotter_splitting_NHC}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
\end{equation}

@@

If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\vt p}_i = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values. These equations can be solved altogether in analytical form, which results in
\begin{subequations}
\begin{align}
&\vt p_i = \vt p_i^0 e^{-\alpha_1^0 t}, \\
&\vt \pi_i = \vt \pi_i^0 e^{-\alpha_1^0 t}, \quad \text{and}\\
&\eta_j = \eta_j^0 + \alpha_j^0 t.
\end{align}
\end{subequations}

On the other hand, if a single term of the last summation was the only term present in Eq.~\ref{eq:iL_NHC}, then $e^{t i\!L_\text{NHC}}$ would behave according to the corresponding equation $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$, with all variables other than $p_{\eta_j}$ kept unchanged. The analytical solution of this single equation is
\begin{equation}
\label{eq:p_eta}
\begin{split}
&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j^0 - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi\left(\alpha_{j+1}^0 t\right) t, \\
\end{split}
\end{equation}
where $\phi(x) = (1-e^{-x})/x$. This is an alternative to the form found in Ref.~\onlinecite{Martyna1996}. We might note here that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^4 \frac{(-1)^n x^n}{(n+1)!}$ whenever  $|x| \leq 10^{-4}$.

Considering the analytical solutions above, one can split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula recursively. The splitting we employ here is
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}} \\
&\times e^{\frac{\Delta t}{2} \left[-\alpha_1 \sum\limits_{i=1}^N \left(\tr{\vt p}_i \frac{\partial}{\partial \vt p_i} + \tr{\vt \pi}_i \frac{\partial}{\partial \vt \pi_i}\right) + \sum\limits_{j=1}^M \alpha_j \frac{\partial}{\partial \eta_j} \right]} \\
&\times \prod_{j=1}^M e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}}.\\
\end{split}
\end{equation}

It is worth noting that analytical solutions exist also for the equations
\begin{align*}
&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i \, \text{ and} \\ 
&\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q_1} \vt \pi_i,
\end{align*}
which are of identical form to Eq.~\ref{eq:p_eta_j}, so they can be evaluated using
\begin{equation}
\label{eq:p}
\begin{split}
{\vt p}_i \leftarrow {\vt p}_i + \left({\vt F}_i - \frac{p_{\eta_1}}{Q_1} {\vt p}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t \, \text{ and}\\
\end{split}
\end{equation}

\begin{equation}
\label{eq:pi}
\begin{split}
{\vt \pi}_i \leftarrow {\vt \pi}_i + \left(2 \tr{\vt \tau} \tr{\mt C} - \frac{p_{\eta_1}}{Q_1} {\vt \pi}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t,
\end{split}
\end{equation}
respectively, with $\phi(x)$ given by Eq.~\ref{eq:phi}. Therefore, one of the aims of this paper is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when also Eqs.~\ref{eq:p} and~\ref{eq:pi} are considered. For that, let as define the modified propagators $ e^{i L^\ast_a} $ and $ e^{i L^\ast_{NHC}} $ as follows:
%\begin{equation}
%\begin{split}
%&i L^\ast_a = i L_a -\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1} \vt p_i^T \diff{}{\vt p_i} - %\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt \pi_i^T  \diff{}{\vt \pi_i} =\\
%&\sum_{i=1}^{N} \left(\tr{\vt F_i} - \frac{p_{\eta_1}}{Q_1}\vt p_i^T \right) \diff{}{\vt p_i} + %\sum_{i=1}^{N} \left(2 \tr{\vt \tau_i} \tr{\mt C_i} -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T\right)  \diff{}{\vt %\pi_i}
%\end{split}
%\end{equation}
%\begin{equation}
%\begin{split}
%i L^\ast_{NHC} = \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} %\frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right] 
%\end{split}
%\end{equation}
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_b^\ast} &= \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} \right] \\
&\times \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \left(2 \tr{\vt \tau}_i \tr{\mt C}_i -  \alpha_1 \tr{\vt \pi}_i \right)  \diff{}{\vt \pi_i}   \right]
\end{split}
\end{equation}

\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } &= \prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
&\times \prod_{j=1}^M \exp\left(-\frac{\Delta t}{2} \alpha_j \diff{}{\eta_j}\right) \\
&\times \prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
\end{split}
\end{equation}
Finally, the complete propagator scheme is
\begin{equation}
\label{eq:trotter_splitting_*}
\begin{split}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast} e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast}.
\end{split}
\end{equation}

In the NVE case, each Liouvillian operator is derived from part of the Hamiltonian, then each factor in the propagator is symplectic and, as a consequence, preserves the phase space volume. Recall that the inverse statement is not necessarily true. Now it is widely recognized the importance of the symplectic property for numerical stability,\cite{Skeel1997} however it is restricted to Hamiltonian systems. For non-Hamiltonian equations of motion the phase space volume is no longer conserved. Nevertheless, there exists an invariant measure on the phase space, and for the the system of equations~\ref{eq:nhc_system} it is given by
\begin{align*}
\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ].
\end{align*}

In this sense, Ezra \cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which are both time  reversible an measure-preserving. In his work, he also pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to more ergodic trajectory behavior. It can be shown that many of the individual operators in both Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*} do not preserve the invariant-measure. Nevertheless, as will be shown below, those numerical schemes successfully achieve the specified temperature and the long-term energy conservation is acceptable. Note that the term that premultiplies the extended phase space volume in the invariant-measure above is a function of the variable $\eta$. Given that the time evolution of the physical variables along with the thermostat momentum does not depend on $\eta$ we conclude that in this case the measure-preserving condition for every individual operator can actually be relaxed.

@@

The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
\[
P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
\]
The first term in the expression above accounts for the ideal gas contribution. Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
\begin{equation}
\begin{split}
\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right. \\ \left. \diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
\end{split}
\end{equation}
Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$. Therefore,
\[
\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \diff{u}{{\vt r}_{ij}} \cdot {\vt r}_{ij}.
\]

@@

\begin{equation}
\label{eq:pressure}
P = \frac{(N-1) k_B T}{V} + \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty}{\vt F}_{ij} \cdot {\vt r}_{ij}  \right\rangle.
\end{equation}
This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs. Thus, an atom located outside the central box will contribute to this virial if it belongs to a body whose center lies in the box, and vice-versa.

\subsection{To do}

Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
\[
W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^{\infty} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ij}.
\]

>>>>

\subsection{Integration of the Canonical Equations of Motion}

As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion. First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components. In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator}. Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as\cite{Martyna1996}
\begin{equation}
\label{eq:trotter_splitting_NHC}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
\end{equation}

@@

If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\eta}_j = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values. This system of equations can be solved analytically, which results in
\begin{subequations}
\begin{align}
&\vt p_i(t) = \vt p_i^0 e^{-\alpha_1 t}, \\
&\vt \pi_i(t) = \vt \pi_i^0 e^{-\alpha_1 t}, \quad \text{and}\\
&\eta_j(t) = \eta_j^0 + \alpha_j t. \label{eq:solution_eta}
\end{align}
\end{subequations}

Otherwise, if a single term $(G_j - \alpha_{j+1} p_{\eta_j}) \partial/\partial p_{\eta_j}$ was the only term present in Eq.~\ref{eq:iL_NHC}, then $e^{t i\!L_\text{NHC}}$ would behave according to the equation $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$, with all variables other than $p_{\eta_j}$ kept unchanged. The analytical solution of this single equation is
\begin{equation}
\label{eq:solution_p_eta}
\begin{split}
&p_{\eta_j}(t) = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1} p_{\eta_j}^0 \right) \phi\left(\alpha_{j+1} t\right) t, \\
\end{split}
\end{equation}
where $\phi(x) = (1-e^{-x})/x$. We might note that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^3 {(-1)^n x^n}/{(n+1)!}$ whenever $|x| \leq 10^{-4}$. The solution above is equivalent to that found in Ref.~\onlinecite{Martyna1996}.

Considering the analytical solutions described above, it is possible to split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula in a recursive way. The splitting we employ here is
\begin{align*}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = &\prod_{j=M}^1 \exp \left[ \frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right] \\
\times \exp \Bigg[ -\frac{\Delta t}{2} &\alpha_1 \sum_{i=1}^N \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j} \Bigg] \\
\times &\prod_{j=1}^M \exp \left[\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].\\
\end{align*}

It is worth noting that an analytical solution can also be derived for the system of equations $\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i$ for all $i \in [1,N]$ if all other variables are constant. These equations are equivalent in form to the one solved above, so that their solutions are
\begin{subequations}
\label{eq:solution_momenta}
\begin{align}
&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \quad \text{and} \label{eq:solution_p} \\
&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t. \label{eq:solution_pi}
\end{align}
\end{subequations}

By evaluating these functions simultaneously for all $i$, we accomplish the effect of a propagator $e^{t i\!L_b^\ast}$, where $i\!L_b^\ast$ is the Liouville operator given by
\[
i\!L_b^\ast = \sum_{i=1}^N \left[ \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} \right].
\]

This is a kind of boost that already incorporates the influence of the thermostat chain. One of the aims of this paper is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when the solution in Eq.~\ref{eq:solution_momenta} is considered. For that, we also define a modified Nos\'e-Hoover chain propagator $e^{i\!L^\ast_\text{NHC}}$, as follows:
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } = &\prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
\times &\exp\left(-\frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j}\right) \\
\times &\prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
\end{split}
\end{equation}

Of course, Eqs.~\ref{eq:solution_eta} and \ref{eq:solution_p_eta} are the ones used to evaluate the propagators above. Finally, the complete scheme is
\begin{equation}
\label{eq:modified_splitting}
\begin{split}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast} e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast}.
\end{split}
\end{equation}

In the microcanonical case, each term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator. As a consequence, the whole propagator preserves the phase-space volume. Recall that the inverse statement is not necessarily true. It is now widely recognized that the symplectic property is important for numerical stability,\cite{Skeel1997} although it is restricted to Hamiltonian systems. For non-Hamiltonian equations of motion, the phase space volume is no longer conserved. Nevertheless, there exists an invariant measure on the phase space, and for the system in Eq.~\ref{eq:nhc_system}, it is given by
\begin{align*}
%\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ].
\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge {\vt r} \wedge {\vt p} \wedge {\vt q} \wedge {\vt \pi} \wedge {\vt \eta} \wedge {\vt p_\eta}
\end{align*}

In this sense, Ezra \cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which are both time-reversible and measure-preserving. In his work, he also pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to trajectories that are more ergodic. It can be shown that many of the individual operators in both Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:modified_splitting} do not preserve the invariant measure. Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. Note that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$. Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.

@@

The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
\[
P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
\]
The first term in the expression above accounts for the ideal gas contribution. Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
\[
%\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right. \\ \left. \diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{{\vt r}_{ij}}{L} \diff{u}{{\vt r}_{ij}} + \diff{{\vt q}_i}{L} \diff{u}{{\vt q}_i} + \diff{{\vt q}_j}{L} \diff{u}{{\vt q}_j} \right).
\]

Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$. Therefore,
\[
\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \tr{\vt r}_{ij} \diff{u}{{\vt r}_{ij}}.
\]

@@

\begin{equation}
\label{eq:pressure}
P = \frac{(N-1) k_B T}{V} + \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty}{\vt F}_{ij} \cdot {\vt r}_{ij}  \right\rangle.
\end{equation}

This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs. Thus, an atom located outside the central box will contribute to this virial if it belongs to a body whose center lies in the box, and vice-versa.

\textbf{To do}: Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
\[
W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^{\infty} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ij}.
\]
